A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-22 text/discussion.Rmd
add PCAWG main paper and evolution refs to disc

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (see **figures 2** and
**supplementary figure S1**) and often are the result of abberations involving
common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
Class-specific evolution must involve the acquisition of combinatorial
disruptions to pathways that may individually be harmful but together produce
more specialized cancer phenotypes and increased fitness (as in **figure 4** and
**tables 1** and **ST2**). Importantly, our work does not distinguish between
the order in which these mutations should accumulate. However, since many of the
genes in the non-specific pathways involve genome stability, it is reasonable to
hypothesize that such mutations promote the acquisition of class specific
defects, and natural selection produces the classes we observed.

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity, both in terms of evolving subclones with different or
passively diverging mutation profiles which then serve as a substrate for futher
selection. Moreover consortium samples are likely to contain contamination from
support tissue and inflammatory and immune cells of the innate and adaptive
immune systems, all potentially harboring cancer supporting mutations
[@bhowmick_review]. We think it will be instructive to explore these ideas in
the context of single cell experiments.

@@

We report that pathway enrichment across all classes (pooled) yielded generally
lower effect sizes and larger p-values than the cluster specific analysis,
suggesting that signal is diluted when the samples are analyzed this way, and
supporting the view that metastasis has class-specific requirements. It should
be noted, however, that in spite of this lack of global enrichment, some of the
cluster specific pathways reflect known metastatic associations, _e.g._ pathways
associated with asparagine and aspartate metabolism in class 10
[@knott_asparagus]. Since the number of metastatic tumors is relatively low this
part of our analysis is likely underpowered and subject to expanded analysis
with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, since (again) we don’t
know the order in which they were acquired in individual tumor samples. The fact
that most metastatic enrichment is cluster-specific and has a tendency to
overlap with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. If this is the case, once we discover the
requirements of each cluster with respect to metastasis we can target them
therapeutically and prevent further adaptation.

>>>>

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (see **figures 2** and
**supplementary figure S1**) and often are the result of aberrations involving
common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
Class-specific evolution must involve the acquisition of combinatorial
disruptions to pathways that may individually be harmful but together produce
more specialized cancer phenotypes and increased fitness (as in **figure 4** and
**tables 1** and **ST2**). Importantly, our work does not distinguish between
the order in which these mutations should accumulate. However, since many of the
genes in the non-specific pathways involve genome stability, it is reasonable to
hypothesize that such mutations promote the acquisition of class specific
defects, and natural selection produces the classes we observed.

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity, both in terms of evolving subclones with different or
passively diverging mutation profiles which then serve as a substrate for futher
selection [@PCAWG-MAIN; @PCAWG-EVO]. Moreover consortium samples are likely to
contain contamination from support tissue and inflammatory and immune cells of
the innate and adaptive immune systems, potentially harboring cancer supporting
mutations [@bhowmick_review]. We think it will be instructive to explore these
ideas in the context of single cell experiments.

@@

We report that pathway enrichment across all classes (pooled) yielded generally
lower effect sizes and larger p-values than the cluster specific analysis,
suggesting that signal is diluted when the samples are analyzed this way, and
supporting the view that metastasis has class-specific requirements. It should
be noted, however, that in spite of this lack of global enrichment, some of the
cluster specific pathways reflect known metastatic associations, _e.g._ pathways
associated with asparagine and aspartate metabolism in class 10
[@knott_asparagus]. Since the number of metastatic tumors is relatively low this
part of our analysis is likely underpowered and subject to expanded analysis
with larger cohorts. Nonetheless, the class-specific metastatic pathway
enrichment we observe (in **table 2**) could be a result of the pathways
directly promoting metastatic phenotypes in these classes, or because they
acquire a stronger or more complete oncogenic phenotype, since (again) we don’t
know the order in which they were acquired in individual tumor samples. The fact
that most metastatic enrichment is cluster-specific and has a tendency to
overlap with cluster-specific pathways from non-metastatic tumors of neighboring
clusters favors the latter interpretation, and suggests a mechanism of
complementarity, in which newly acquired mutations result in similar classes
converging on a single deadly phenotype that captures most or all of the
critical features of end-stage cancer. This is consistent with the observation
that tumors increase diversity of driver mutations in later stages [@PCAWG-EVO].
When we discover the requirements of each cluster with respect to metastasis we
can target them therapeutically and prevent further adaptation.
